---
title: "Nominations analysis"
author: 
  - name: Max H. Feinleib
    orcid: 0009-0002-9604-3533
    email: maxfeinleib2024@u.northwestern.edu
    affiliation:
      - name: Northwestern University
        city: Evanston
        state: IL
        url: https://www.northwestern.edu
bibliography: "full_bib.json"
csl: apa.csl
toc: true
number-sections: true
---

```{r packages}
# packages used
library(dplyr)
library(lubridate)
library(stringr)

library(filibustr)
```


## Introduction
To test the effect of real-life filibuster reforms, I measure the difference in supporting coalitions on presidential nominations before and after the nuclear options in 2013 and 2017.

## Data
I use presidential nominations for executive-branch and judicial-branch positions. I divide nomination votes according to whether they came before or after the nuclear option:
* November 21, 2013 for executive-branch and non-Supreme Court judicial-branch nominees
* April 6, 2017 for Supreme Court nominees
```{r data import}
# starting in 1989, when data became complete
nominations_data <- get_voteview_rollcall_votes(chamber = "s", congress = 101:117) |> 
  filter(str_starts(bill_number, "PN"))

# presidents data for their ideologies
presidents <- get_voteview_members(chamber = "s", congress = 101:117) |> 
  filter(chamber == "President",
         !(congress == 117 & str_starts(bioname, "TRUMP")))

parties <- get_voteview_parties(chamber = "s", congress = 101:117)

nuclear_2013 <- as.Date("2013-11-21")
nuclear_2017 <- as.Date("2017-04-06")
```

Transformation: 
* Filtering to the confirmation vote
* Determining which votes came after the nuclear option
```{r data transformation}
party_sizes <- parties |> 
  filter(party_name != "Independent", 
         # filter out Trump for 117th
         !(congress == 117 & chamber == "President" & party_name == "Republican")) |> 
  mutate(pres_party = first(party_name), .by = congress, .after = congress) |> 
  # just Senate parties
  filter(chamber != "President") |> 
  mutate(pct_members = n_members / sum(n_members) * 100, .by = congress, .after = n_members) |>
  # just the President's party
  filter(pres_party == party_name) |> 
  select(congress, pres_party, pres_party_size = pct_members)

nominations <- nominations_data |> 
  filter(vote_question == "On the Nomination") |> 
  mutate(
    # SCOTUS nominations
    scotus = str_detect(vote_desc, "(Associate|Chief) Justice"),
    # votes after the nuclear option
    post_nuclear = (
      (date > nuclear_2017 | (congress == 115 & rollnumber == 110)) 
      | (!scotus & (date > nuclear_2013 | (congress == 113 & rollnumber == 244)))
    ),
    vote_pct = yea_count / (yea_count + nay_count) * 100,
    yrs_since_89 = interval(as.Date("1989-01-01"), date) / years(1)
  ) |> 
  inner_join(party_sizes)
```

```{r models}
# confirmation vote percentages
plot(nominations$date, nominations$vote_pct)

# Model 1: by Congress
formula_congress <- reformulate(response = "vote_pct",
                                termlabels = c("post_nuclear", "scotus", 
                                               "I(congress - 100)",
                                               "pres_party_size"))
pred_vote <- lm(formula_congress,
                data = nominations)
summary(pred_vote)



# Model 2: by time since 1989
pred_vote_dates <- lm(vote_pct ~ post_nuclear + scotus + yrs_since_89 + I(yrs_since_89 ^ 2) + I(yrs_since_89 ^ 3) + pres_party_size,
   data = nominations)
summary(pred_vote_dates)

# diagnostic plots
plot(nominations$date, pred_vote$fitted.values)
plot(nominations$date, pred_vote_dates$fitted.values)

plot(pred_vote)[1]
plot(pred_vote_dates)

nominations |> filter(congress >= 107)

# Model 3: ridge regression
ridge_model <- glmnet::cv.glmnet(y = nominations$vote_pct, 
                                 x = as.matrix(nominations |> 
                                                 select(post_nuclear, scotus, 
                                                        yrs_since_89, pres_party_size)),
                                 alpha = 0, lambda = 10^seq(2, -2, by = -.1))
summary(ridge_model)
ridge_model$lambda.min
plot(ridge_model)
coef(ridge_model, s = "lambda.min")
```

